---
title: "spectracle -- OMIP-109"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
format:
  html:
    df-print: paged
---

```{r}
#| label: setup
library(spectracle)
```

# Objective

Derive spectra using OMIP-109 source data.

# Source Data

Source data as presented in [OMIP-109](http://www.ncbi.nlm.nih.gov/pubmed/39466962) is downloaded from its public repository:

* [figshare -- OMIP-109](https://doi.org/10.6084/m9.figshare.25699221)

```{r}
#| label: source-data
#| code-fold: true
#| code-summary: "Click to see how OMIP-109 source data was obtained."

## manually download or use this script

## source directory; this path is user-dependent;
## the author of this article previously downloaded these source files
dir.source <- "W:/OMIP-109_flowstate/data_source/OMIP-XXX 45-Color Experiment 1/"


## figshare name: 45-Color Full Spectrum Flow Cytometry Panel
## Experiment ($PROJ) name: OMIP-XXX 45-Color Experiment 1
## Size: 1.74 GB
## MD5 checksum: 623c49ca26c4584a3bcc44b6a3fa5843
## download link: https://ndownloader.figshare.com/files/45873810
## Citation: Jaimes, Maria (2024). 45-Color Full Spectrum Flow Cytometry Panel. 
## figshare. Dataset. https://doi.org/10.6084/m9.figshare.25699221.v1


## test for existence and integrity of source files
if(dir.exists(dir.source)){
  digests <- sapply(list.files(dir.source, full.names = T, recursive = T),
                    digest::digest,
                    file = TRUE)
  digest <- digest::digest(as.vector(digests))
  if(digest != "4d9672f8142a8cde3aa7dfdf3833acc7"){
    message("Digest of source files (.fcs) does not match historic value;\ndownloading source files...")
    download.source <- TRUE
  }else{
    download.source <- FALSE
  }
}else{
  download.source <- TRUE
}
if(download.source){
  temp <- tempfile(tmpdir = "data_temp/", fileext = ".zip")
  zip.link <- "https://ndownloader.figshare.com/files/45873810"
  curl::curl_download(
    url = zip.link,
    destfile = temp,
    quiet = FALSE,
    mode = "wb"
  )
  ## check the hash
  if (tools::md5sum(temp) != "623c49ca26c4584a3bcc44b6a3fa5843") {
    stop("MD5 checksum (hash) of the downloaded .zip does not match the historic value.")
  }
  ## files contained in the .zip: 100 files;
  ## Raw, Unmixed, and SpectroFlo software files (.Expt, .WTML, and .UST)
  utils::unzip(temp, list = TRUE)
  ## unzip to "./data_temp"
  utils::unzip(temp, files = NULL, exdir = "./data_temp/")
  ## get the original 'Project' ('$PROJ') name to reconstruct directory structure
  proj <- "./data_temp/Raw/Reference Group/Unstained (Cells).fcs" |>
    flowstate:::readFCStext() |>
    flowstate:::keywords.to.data.table() |>
    _[][['$PROJ']]
  ## R scripts directory
  dir.analysis <- file.path("R", proj)
  if (!dir.exists(dir.analysis))
    dir.create(dir.analysis)
  ## create the source directory
  dir.out <- file.path("data_source/", proj)
  if (!dir.exists(dir.out))
    dir.create(dir.out)
  ## move the source .fcs files; Raw and Unmixed
  fcs.to.move <- list.files(
    "data_temp/",
    full.names = T,
    pattern = ".fcs",
    recursive = T
  )
  dirnames <- dirname(fcs.to.move) |> unique()
  sapply(dirnames, function(i) {
    dir.out.sub <- sub("data_temp", dir.out, i)
    if (!dir.exists(dir.out.sub))
      dir.create(dir.out.sub, recursive = T)
  }) |> invisible()
  file.rename(from = fcs.to.move,
              to = sub("data_temp", dir.out, fcs.to.move)) |> invisible()
  ## move the SpectroFlo experiment files
  dir.out.expt <- file.path(dir.out, "SpectroFlo_experiment")
  if (!dir.exists(dir.out.expt))
    dir.create(dir.out.expt)
  files.to.move <- list.files("data_temp/", full.names = T, pattern = "Expt|UST|WTML")
  file.rename(from = files.to.move,
              to = sub("data_temp", dir.out.expt, files.to.move)) |> invisible()
  ## cleanup
  unlink(temp)
  unlink("data_temp/Raw/", recursive = T)
  unlink("data_temp/Unmixed//", recursive = T)
}
##
```

## Source Files -- Raw Reference Controls (Cells)

```{r}
#| label: source-data-raw-reference-files
#| code-fold: true
#| code-summary: "'./Raw/Reference Group' file paths"

dir.source.raw <- grep("Raw.*Ref", list.dirs(dir.source), value = T)
raw.reference.controls <- list.files(
  dir.source.raw, 
  full.names = T, 
  pattern = ".fcs"
)
##
data.frame(basename(raw.reference.controls))
```

# Derive Spectra

`spectracle()` is fully automated and by design has limited function arguments; simply pass the directory where the source files are contained.

```{r}
#| label: spectracle

spectra <- spectracle(dir.source.raw)
```

## Spectra -- OMIP-109

`spectracle()` returns a standardized `data.table::data.table()` containing relevant metadata:

* deriving function
* date-of-acquisition
* cytometer, serial number, software
* sample identifier (sample name, fluor, marker/stain, peak detector, laser)
* derived, normalized [0,1] spectra
* hash summaries for each spectra

```{r}
#| label: spectra-omip109

spectra[]
```

## Spectra -- Hashes

Derived using a fully data-driven/automated process, the resultant spectra should reproduce given the same source data and function version.  A summary digest is generated from the individual hashes: 

* Hash (md5) summary: `{r} spectra[, digest::digest(hash.md5)]`

```{r}
#| label: spectra-omip109-hashes

spectra[, .(N, S, detector, hash.md5)]
```

## Spectra -- Traces

```{r, results='asis', echo=FALSE}
#| label: plot-traces
#| 
lasers <- spectra[, c(levels(laser), "AF")]
spectra.list <- split(spectra[N != "AF"], by = 'laser')
spectra.list[['AF']] <- spectra[N == "AF"]
## each unique laser + AF as a tabset
cat("::: {.panel-tabset}\n")
for(.laser in lasers){
  cat("\n###", .laser, "\n")
  ## laser-specific trace plots as a tabset
  cat("::: {.panel-tabset}\n")
  for(id in spectra.list[[.laser]][, as.character(sample.id)]){#.names
    cat("\n#### ", id, "\n")
    plot_trace(spectra.list[[.laser]][sample.id == id])
    cat("\n")
  }
  cat(":::\n")
  cat("\n")
}
cat(":::\n")
```
